A weakly structured stem for human origins in Africa

### 2023-01-08 paper.tex
edits to main text

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}, but these
approaches are more challenging to interpret and fit to data.
However, new population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent articles
have shown that introducing an archaic hominin ghost population contributing to 
African populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models, mostly in Western Africa
\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,Lorente-Galdos2019-vz,Durvasula2020-td},
%Plagnol: YRI 
%Here we present a new method for addressing whether archaic human groups contributed to the modern gene pool (called ancient admixture),
%we find strong evidence for ancient admixture in both a European and a West African population
%
%Hammer: Mandenka, Biaka, and San. Title:
%"Genetic evidence for archaic admixture in Africa"
%Abstract:"allow us to infer that contemporary African populations contain a small proportion of genetic material (≈2%) that introgressed ≈35 kya from an archaic population that split from the ancestors of anatomically modern humans ≈700 kya."
%
%Hey: Strongest signal in Yoruba, Baka
% Abstract: "consistent with human/archaic divergence."
%
%Ragsdale: YRI
%abstract: We estimate that an unidentified, deeply diverged population admixed with modern humans within Africa both before and after the split of African and Eurasian populations, contributing 4 − 8% genetic ancestry to individuals in world-wide populations.
%
%Lorente Galdos: Khoesan, Mbuti, W. AFR
%title:"reveals archaic gene flow"
%
%Durvasula: West African
%title: "Recovering signals of ghost archaic introgression"
%Abstract:"Our analyses of site frequency spectra indicate that these populations derive 2 to 19% of their genetic ancestry from an archaic population that diverged before the split of Neanderthals and modern humans." 
but also in southern \cite{Hammer2011-bx,Lorente-Galdos2019-vz} 
and central African \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz,Hey2018-pw}.
populations.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence \citep[e.g.,][]{Hsieh2016-gk}. However, these prior
articles share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African or Bantu-speaking ancestry
(Figure~\ref{fig:diversity}C). 
While ancient DNA from Eurasia has helped us understand early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.5\textwidth]{figures/proposed-models.png}
    \caption{
        \textbf{Proposed conceptual models of early human history in Africa.}
        These models have been designed to translate models from the 
        paleoanthropological literature into genetically testable demographic
        models \citep{Henn2018-rf}.
        We used these conceptual models as starting points to build detailed
        parameterized demographic models (SI Section~\ref{modelspec})
        that were then fit to genetic data. 
    }
    \label{fig:proposed-models}
\end{figure}

@@

We began with a model of geographic expansion from a single ancestral,
unstructured source followed by migration between populations, without allowing
for contribution from an African archaic hominin lineage or population structure prior
to the expansion (Figure~\ref{fig:proposed-models}A or Figure~\ref{fig:proposed-models}D). As expected
\citep{Ragsdale2019-nt}, this first model was a poor fit to the data
qualitatively (Figure~\ref{fig:supp-single-origin-fits}) and quantitatively
(log-likelihood ($LL$) $\approx -189,300$, Table~\ref{tab:supp-single-origin}).
We next explored a suite of parameterized models in which population structure predates the
differentiation of contemporary groups (Supporting Information (SI) Section~\ref{modelspec}). 
Depending on the parameters, these encompassed models allowing for ancestral
reticulation (such as fragmentation-and-coalescence or meta-population models,
Figure~\ref{fig:proposed-models}B), archaic hominin admixture
(Figure~\ref{fig:proposed-models}C), and African multi-regionalism
(Figure~\ref{fig:proposed-models}D).

@@

We quantify migration rates among populations after their divergence
$\approx120ka$. Prior to agripastoralist expansion 5ka, migration between the ancestors
of the Nama and other groups is an order of magnitude weaker than that observed
between between western and eastern Africans (Table \ref{tab:migration-rates}).
All models infer relatively high gene flow between Ethiopia and western Africa
($m\approx2\times10^{-4}$, the constant proportion of immigrant lineages per
generation over 60ka). We further find that Back-to-Africa gene flow at the
beginning of the Holocene primarily affected the ancestors of the Ethiopian
agricultural populations \citep{Molinaro2019-rf}, comprising over half of their
genetic ancestry, estimated to be approximately 65\%. We observe significant gene flow
from the Amhara and Oromo into the Nama, a signal which is likely a proxy for
the movement of eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 ya. While this gene flow is not apparent from the ADMIXTURE
plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped into the
Khoe-San component which has drifted appreciably from its ancestral eastern
African source. Colonial period admixture from Europeans into the Nama was
estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration between stems
($LL \approx -101,600$ vs. $-107,700$ in the merger model and
$LL \approx -115,300$ vs. $-126,500$ in the continuous migration model).
With continuous migration between stems, population
structure extends back to 1.1--1.4Ma (Table~\ref{tab:migration-rates}).
Migration between the stems in these models is moderate, with a fraction of
migrant lineages each generation estimated as
$m=6.3\times10^{-5}-1.3\times10^{-4}$. For comparison, this is similar
to inferred migration rates between connected contemporary populations over the
past 50ka (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic hominin admixture models (Figure~\ref{fig:proposed-models}C) as
the early branches remain closely related, and each branch contributes large
amounts to all contemporary populations (Figure~\ref{fig:predictions}). 
Because of this relatedness, only $1\%$ to $4\%$ of genetic differentiation
among contemporary populations can be traced back
to this early population structure (SI Section~\ref{sec:f4})

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups. 
We observe a sharp bottleneck in Stem 1 down to $N_e=100$ after the split of
the Neanderthal branch. This represents the lower bound allowed in our
optimization (i.e. an $N_e$ of 100), although the size of this
bottleneck is poorly constrained (95\% confidence interval 100--851). After a long period of
exchange with Stem 2, Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S''
478ka. The timing of this divergence was also poorly constrained (95\% CI 276–-478ka).
These populations evolve independently until 119ka
(101--125ka) when Stem 1S and Stem 2 combine to form the ancestors of the Nama, with
proportions 30\%, 70\% respectively. Similarly, Stem 1E and Stem 2 combine in
equal proportions (50\% each) to form the ancestors of the western Africans and
eastern Africans (and thus also all individuals who later disperse during the
Out of Africa event). Finally, the Mende receive a large additional pulse of
gene flow from Stem 2, replacing 19\% of their population 25ka (22--26ka).
The later Stem 2 contribution to the western African Mende resulted in
better model fits ($\Delta LL \approx 60,000$).
Speculatively, this may indicate that an ancestral Stem 2 population occupied western or
central Africa, broadly speaking. The differing proportions in the Nama and
eastern Africans may also indicate geographic separation of Stem 1S in southern
Africa and Stem 1E in eastern Africa.

\begin{figure}[t!]
    \centering
    \includegraphics{figures/predictions.pdf}
    \caption{
        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains contemporary structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled contemporary
        populations that aligns with drift between past populations
        (here Nama and Mende, see SI Section~S\ref{sec:f4}
        for details and additional comparisons in
        Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.
        Most present differentiation dates back to the last 100ka.
    }
    \label{fig:predictions}
\end{figure}
To assess robustness of the inferred models to analysis and reference
population choices, SI Sections \ref{sec:validation_strategy}
and \ref{sec:validation_results} present re-analyses changing the European and
West African populations, as well as the recombination maps, filtering
strategies and parameter optimization strategies. While we find some
differences in the inferred parameters (see in particular Sections
\ref{sec:replacing} and \ref{sec:change-dates}), the best fit model across all
reanalyses are quantitatively consistent.

>>>>

Genetic models have been hampered in their contribution to this discussion
because they primarily assume (or, at least, have been tested under) a
tree-like model of isolation-with-migration. Alternative theoretical scenarios
have been proposed, such as stepping stone models \citep{Arredondo2021-qa} or
population coalescence and fragmentation \citep{Scerri2019-xg}, but these
approaches are more challenging to interpret and fit to data.
However, new population genetic tools now allow for inference involving tens to
hundreds of genomes from multiple populations and greater complexity
\citep{Kamm2020-vn,Ragsdale2019-nt,Speidel2019-nj}. Inspired by evidence for
Neanderthal admixture with humans in Eurasia, several recent studies
have shown that introducing an archaic hominin ghost population contributing to 
African populations in the period surrounding the Out-of-Africa migration event
substantially improves the description of genetic data relative to
single-origin models, mostly in Western Africa
\citep{Plagnol2006-lt,Hammer2011-bx,Hey2018-pw,Ragsdale2019-nt,
Lorente-Galdos2019-vz,Durvasula2020-td},
%Plagnol: YRI 
%Here we present a new method for addressing whether archaic human groups contributed to the modern gene pool (called ancient admixture),
%we find strong evidence for ancient admixture in both a European and a West African population
%
%Hammer: Mandenka, Biaka, and San. Title:
%"Genetic evidence for archaic admixture in Africa"
%Abstract:"allow us to infer that contemporary African populations contain a small proportion of genetic material (≈2%) that introgressed ≈35 kya from an archaic population that split from the ancestors of anatomically modern humans ≈700 kya."
%
%Hey: Strongest signal in Yoruba, Baka
% Abstract: "consistent with human/archaic divergence."
%
%Ragsdale: YRI
%abstract: We estimate that an unidentified, deeply diverged population admixed with modern humans within Africa both before and after the split of African and Eurasian populations, contributing 4 − 8% genetic ancestry to individuals in world-wide populations.
%
%Lorente Galdos: Khoesan, Mbuti, W. AFR
%title:"reveals archaic gene flow"
%
%Durvasula: West African
%title: "Recovering signals of ghost archaic introgression"
%Abstract:"Our analyses of site frequency spectra indicate that these populations derive 2 to 19% of their genetic ancestry from an archaic population that diverged before the split of Neanderthals and modern humans." 
but also in southern \cite{Hammer2011-bx,Lorente-Galdos2019-vz} 
and central African \cite{Hsieh2016-gk, Hammer2011-bx,Lorente-Galdos2019-vz,Hey2018-pw}
populations.
This has driven speculation about the geographic range of this ghost
population, possible links to specific fossils, and the possibility of
finding ancient DNA evidence\citep{Hsieh2016-gk}. However, these prior
studies share two weaknesses. First, they only contrast a single-origin model
with an archaic hominin admixture model, leaving out other plausible models
\citep{Henn2018-rf} (Figure~\ref{fig:proposed-models}).
Second, they
focus on a small subset of African diversity, either because of small sample
sizes (2-5 genomes) or because they rely on 1000 Genomes data which only
recruited populations of recent West African or Bantu-speaking ancestry
(Figure~\ref{fig:diversity}C). 
While ancient DNA from Eurasia has helped to clarify early human history
outside of Africa, there is no comparably ancient DNA to elucidate early history
in Africa \citep{Lipson2022-xf}.

@@

\begin{figure}[ht]
    \centering
    \includegraphics[width=0.5\textwidth]{figures/proposed-models.png}
    \caption{
        \textbf{Proposed conceptual models of early human history in Africa.}
        These models have been designed to translate models from the 
        paleoanthropological literature into genetically testable demographic
        models \citep{Henn2018-rf}.
        We used these conceptual models as starting points to build detailed
        parameterized demographic models (SI Section~\ref{sec:modelspec})
        that were then fit to genetic data. 
    }
    \label{fig:proposed-models}
\end{figure}

@@

We began with a model of geographic expansion from a single ancestral,
unstructured source followed by migration between populations, without allowing
for contribution from an African archaic hominin lineage or population
structure prior to the expansion (Figure~\ref{fig:proposed-models}A or
Figure~\ref{fig:proposed-models}D). As expected \citep{Ragsdale2019-nt}, this
first model was a poor fit to the data qualitatively
(Figure~\ref{fig:supp-single-origin-fits}) and quantitatively (log-likelihood
($LL$) $\approx -189,300$, Table~\ref{tab:supp-single-origin}).  We next
explored a suite of parameterized models in which population structure predates
the differentiation of contemporary groups (Supporting Information (SI)
Section~\ref{sec:modelspec}).  Depending on the parameters, these encompassed
models allowing for ancestral reticulation (such as
fragmentation-and-coalescence or meta-population models,
Figure~\ref{fig:proposed-models}B), archaic hominin admixture
(Figure~\ref{fig:proposed-models}C), and African multi-regionalism
(Figure~\ref{fig:proposed-models}D).

@@

We quantify migration rates among populations after their divergence
$\sim$120ka. Prior to agripastoralist expansion 5ka, migration between the
ancestors of the Nama and other groups is an order of magnitude weaker than
that observed between between western and eastern Africans (Table
\ref{tab:migration-rates}). All models infer relatively high gene flow between
Ethiopia and western Africa ($m\approx2\times10^{-4}$, the constant proportion
of immigrant lineages per generation over 60ka). We further find that
Back-to-Africa gene flow at the beginning of the Holocene primarily affected
the ancestors of the Ethiopian agricultural populations
\citep{Molinaro2019-rf}, comprising over half of their genetic ancestry,
estimated to be approximately 65\%. We observe significant gene flow from the
Amhara and Oromo into the Nama, a signal which is likely a proxy for the
movement of eastern African caprid and cattle pastoralists
\citep{Henn2008-xo,Breton2014-xb}, here estimated to constitute a 25\% ancestry
contribution 2,000 years ago. While this gene flow is not apparent from the
ADMIXTURE plot (Figure~\ref{fig:diversity}), the ancestry is likely grouped
into the Khoe-San component which has drifted appreciably from its ancestral
eastern African source. Colonial period admixture from Europeans into the Nama
was estimated at 15\%, similar to proportions suggested by \texttt{ADMIXTURE}
(Figure~\ref{fig:diversity}).

@@

Allowing for continuous migration between the stem populations substantially
improved the fits relative to zero migration between stems ($LL \approx
-101,600$ vs. $-107,700$ in the merger model and $LL \approx -115,300$ vs.
$-126,500$ in the continuous migration model).  With continuous migration
between stems, population structure extends back to up to 1.5Ma
(Table~\ref{tab:migration-rates}).  Migration between the stems in these models
is moderate, with a fraction of migrant lineages each generation estimated as
$m=6.3\times10^{-5}-1.3\times10^{-4}$. For comparison, this is similar to
inferred migration rates between connected contemporary populations over the
past 50ka (Table~\ref{tab:migration-rates}). This ongoing (or at least,
periodic) gene flow qualitatively distinguishes these models from previously
proposed archaic hominin admixture models (Figure~\ref{fig:proposed-models}C)
as the early branches remain closely related, and each branch contributes large
amounts to all contemporary populations (Figure~\ref{fig:predictions}).
Because of this relatedness, only $1\%$ to $4\%$ of genetic differentiation
among contemporary populations can be traced back to this early population
structure (SI Section~\ref{sec:f4})

@@

In contrast, under the multiple-merger model, stem populations merge with
varying proportions to form the different contemporary groups.  We observe a
sharp bottleneck in Stem 1 down to $N_e=100$ after the split of the Neanderthal
branch. This represents the lower bound allowed in our optimization (i.e. an
$N_e$ of 100), although the size of this bottleneck is poorly constrained (95\%
confidence interval 100--851). After a long period of exchange with Stem 2,
Stem 1 then fractures into ``Stem 1E'' and ``Stem 1S'' 478ka. The timing of
this divergence was also poorly constrained (95\% CI 276–-478ka).  These
populations evolve independently until 119ka (101--125ka) when Stem 1S and Stem
2 combine to form the ancestors of the Nama, with proportions 30\%, 70\%
respectively. Similarly, Stem 1E and Stem 2 combine in equal proportions (50\%
each) to form the ancestors of the western Africans and eastern Africans (and
thus also all individuals who later disperse during the Out of Africa event).
Finally, the Mende receive a large additional pulse of gene flow from Stem 2,
replacing 19\% (18--21\%) of their population 25ka (22--26ka). The later Stem 2
contribution to the western African Mende resulted in better model fits
($\Delta LL \approx 60,000$).  Speculatively, this may indicate that an
ancestral Stem 2 population occupied western or central Africa, broadly
speaking. The differing proportions in the Nama and eastern Africans may also
indicate geographic separation of Stem 1S in southern Africa and Stem 1E in
eastern Africa.

\begin{figure}[t!]
    \centering
    \includegraphics{figures/predictions.pdf}
    \caption{
        \textbf{Structure among stems is weak and present-day structure is mostly recent.}
        From the best fit models of our two parameterizations (A and B:
        continuous migration, C and D: merger with stem migration),
        we predicted pairwise differences $H_{i,j}$ between individuals
        sampled from populations $i$ and $j$ existing at time $t$ (A and C).
        (B and D) To understand how drift between stems explains contemporary structure, 
        we also computed the proportion $\alpha^2$ of drift between pairs of
        sampled contemporary
        populations that aligns with drift between past populations
        (here Nama and Mende, see SI Section~S\ref{sec:f4}
        for details and additional comparisons in
        Figures~\ref{fig:supp-f4s-single-origin}--\ref{fig:supp-f4s-merger-with-stem-migration}).
        Both models infer deep population structure with modest contributions to
        contemporary genetic differentiation.
        Most present differentiation dates back to the last 100ka.
    }
    \label{fig:predictions}
\end{figure}

To assess robustness of the inferred models to analysis and reference
population choices, SI Sections \ref{sec:validation_strategy}
and \ref{sec:validation_results} present re-analyses changing the European and
West African populations, as well as the recombination maps, filtering
strategies and parameter optimization strategies. While we find some
differences in the inferred parameters (see in particular Sections
\ref{sec:replacing} and \ref{sec:change-dates}), the best fit model across all
reanalyses are quantitatively consistent.
